---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
set.seed(1)
```

# sondage

<!-- badges: start -->
[![R-CMD-check](https://gitlab.com/dickoa/sondage/badges/main/pipeline.svg)](https://gitlab.com/dickoa/sondage/-/pipelines)
[![Codecov test coverage](https://codecov.io/gl/dickoa/sondage/branch/main/graph/badge.svg)](https://app.codecov.io/gl/dickoa/sondage?branch=main)
<!-- badges: end -->

Fast survey sampling algorithms for `R`. Sampling functions return **design objects** with generics for extracting inclusion probabilities, joint inclusion probabilities, and variance estimation quantities.

For without-replacement designs, the stored `pik` vector is the design-defining target inclusion probability vector. For methods with exact first-order guarantees, this equals the true first-order inclusion probabilities. For order-sampling methods such as `sps` and `pareto`, the stored vector remains the target `pik`, while the true finite-population first-order inclusion probabilities are only approximately equal to that target.

## Installation

```{r, eval = FALSE}
# From GitLab
pak::pkg_install("gitlab::dickoa/sondage")
```

## Usage

```{r}
library(sondage)

# Use built-in US state data
data(state)
states <- as.data.frame(state.x77)

# Compute inclusion probabilities from population size
pik <- inclusion_prob(states$Population, n = 10)

# Draw a sample (Conditional Poisson Sampling)
s <- unequal_prob_wor(pik, method = "cps")
states[s$sample, ]
```

```{r}
# Joint inclusion probabilities for variance estimation
pikl  <- joint_inclusion_prob(s)
delta <- sampling_cov(s)                # pi_ij - pi_i * pi_j
chk   <- sampling_cov(s, weighted = TRUE) # 1 - pi_i * pi_j / pi_ij
```

```{r}
# Equal probability sampling
s <- equal_prob_wor(nrow(states), 10)
states[s$sample, ]
```

```{r}
# PPS with minimum replacement (Chromy)
hits <- expected_hits(states$Population, n = 10)
s <- unequal_prob_wr(hits, method = "chromy")
```

```{r}
# Balanced sampling (cube method)
pik <- inclusion_prob(states$Population, n = 10)
x <- matrix(states$Income)
s_bal <- balanced_wor(pik, aux = x)
s_bal
```

```{r}
# Controlled selection e.g to keep the number of sampled states per region
# within the integers adjacent to its expectation
B <- sapply(levels(state.region), function(g) as.double(state.region == g))
S <- colSums(B * pik)
s_ctrl <- balanced_wor(
  pik,
  bounds = list(B = B, lower = floor(S), upper = ceiling(S))
)
table(state.region[s_ctrl$sample])
```

```{r}
# Batch sampling for simulations (design object with matrix $sample)
sim <- unequal_prob_wor(pik, method = "cps", nrep = 1000)
dim(sim$sample)   # 10 x 1000
inclusion_prob(sim) # generics still work
```

## Sampling functions

**Equal probability without replacement** (`equal_prob_wor`):

- `equal_prob_wor(N, n, method = "srs")` - Simple random sampling
- `equal_prob_wor(N, n, method = "systematic")` - Systematic sampling
- `equal_prob_wor(N, n, method = "bernoulli")` - Bernoulli sampling (random size)

**Equal probability with replacement** (`equal_prob_wr`):

- `equal_prob_wr(N, n, method = "srs")` - Simple random sampling with replacement

**Unequal probability without replacement** (`unequal_prob_wor`):

- `unequal_prob_wor(pik, method = "cps")` - Conditional Poisson / maximum entropy
- `unequal_prob_wor(pik, method = "sampford")` - Exact Sampford PPS with exact joint probabilities
- `unequal_prob_wor(pik, method = "brewer")` - Brewer's method
- `unequal_prob_wor(pik, method = "systematic")` - Systematic PPS
- `unequal_prob_wor(pik, method = "poisson")` - Poisson sampling (random size)
- `unequal_prob_wor(pik, method = "sps")` - Sequential Poisson sampling (order sampling)
- `unequal_prob_wor(pik, method = "pareto")` - Pareto sampling (order sampling)

**Unequal probability with replacement** (`unequal_prob_wr`):

- `unequal_prob_wr(hits, method = "chromy")` - PPS with minimum replacement
- `unequal_prob_wr(hits, method = "multinomial")` - Multinomial PPS

**Balanced sampling without replacement** (`balanced_wor`):

- `balanced_wor(pik, aux, method = "cube")` - Cube method (Deville & Tillé, 2004)
- `balanced_wor(pik, aux, strata, method = "cube")` - Stratified cube (Chauvet, 2009)
- `balanced_wor(pik, bounds = list(B, lower, upper))` - Cube with inequality
  constraints (Tripet & Tillé, 2026) providing controlled matrix rounding, minimum group sizes
- `balanced_wor(pik, spread, method = "lpm2")` - Spatially balanced, well-spread
  sampling with the local pivotal method 2 (Grafström, Lundström & Schelin, 2012)
- `balanced_wor(pik, spread, method = "scps")` - Spatially correlated Poisson
  sampling with Grafström's (2012) maximal-weight strategy

## Design queries

- `inclusion_prob(x, n)` - Compute inclusion probabilities from size measures
- `inclusion_prob(s)` - Extract the stored design-defining `pik` vector from a WOR design
- `expected_hits(x, n)` - Compute expected hits from size measures
- `expected_hits(s)` - Extract expected hits from a WR design
- `joint_inclusion_prob(s)` - Joint inclusion probabilities (WOR)
- `joint_inclusion_prob(s, sampled_only = TRUE)` - n x n submatrix for sampled units only (scales to large N)
- `joint_expected_hits(s)` - Pairwise expectations E(n_i n_j) (WR)
- `joint_expected_hits(s, sampled_only = TRUE)` - Submatrix for selected units only
- `sampling_cov(s)` - Sampling covariance matrix
- `sampling_cov(s, weighted = TRUE)` - Check quantities for SYG variance estimator
- `sampling_cov(s, sampled_only = TRUE)` - Covariance for sampled units only

Joint and covariance matrices preserve population-unit names. When the
probability vector is unnamed, sampled-only matrices use population indices as
row and column names so their units remain identifiable.

## Method comparison

| Method        | Dispatcher         | Fixed n | Exact marginals† | Exact pi_ij        | PRN |
|---------------|--------------------|---------|------------------|--------------------|-----|
| `srs`         | `equal_prob_wor`   | yes     | yes              | yes                | no  |
| `systematic`  | `equal_prob_wor`   | yes     | yes              | yes                | no  |
| `bernoulli`   | `equal_prob_wor`   | no      | yes              | yes (independent)  | yes |
| `srs`         | `equal_prob_wr`    | yes     | yes              | yes (analytic)     | no  |
| `cps`         | `unequal_prob_wor` | yes     | yes              | yes                | no  |
| `sampford`    | `unequal_prob_wor` | yes     | yes              | yes                | no  |
| `brewer`      | `unequal_prob_wor` | yes     | yes              | approx (HE)        | no  |
| `systematic`  | `unequal_prob_wor` | yes     | yes              | yes (some = 0)     | no  |
| `poisson`     | `unequal_prob_wor` | no      | yes              | yes (independent)  | yes |
| `sps`         | `unequal_prob_wor` | yes     | target only*     | approx (HE)**      | yes |
| `pareto`      | `unequal_prob_wor` | yes     | target only*     | approx (HE)**      | yes |
| `multinomial` | `unequal_prob_wr`  | yes     | yes              | yes (analytic)     | no  |
| `chromy`      | `unequal_prob_wr`  | yes     | yes              | simulated          | no  |
| `cube`        | `balanced_wor`     | yes     | yes              | approx (HE)        | no  |
| `lpm2`        | `balanced_wor`     | yes     | yes              | not available      | no  |
| `scps`        | `balanced_wor`     | yes     | yes              | not available      | no  |

†For WOR methods, design marginals are first-order inclusion probabilities `\pi_k`. For WR methods, design marginals are expected hits `E(N_k)`.

\*For `sps` and `pareto`, `inclusion_prob(s)` returns the stored design-defining target vector; the true finite-population first-order inclusion probabilities are only approximately equal to that target, with the discrepancy vanishing asymptotically.

\**For `sps` and `pareto`, the high-entropy approximation is built from the
stored target `pik` vector. HE = high-entropy approximation.

## Choosing a method

- Use `cps` when exactness matters more than speed: it is the maximum-entropy fixed-size unequal-probability design, with exact first and second-order inclusion probabilities.
- Use `brewer` when you want exact first-order inclusion probabilities with lower computational cost than `cps`, and can work with approximate second-order quantities in `sondage`.
- Use `sampford` for the SAS-familiar fixed-size PPS design with exact first- and second-order inclusion probabilities. Its C kernel uses complement sampling and a bounded-rejection/non-rejective hybrid.
- Use `systematic` when very fast sampling and ordering or implicit stratification are central, and structural zeros in some joint inclusion probabilities are acceptable.
- Use `poisson` when a random sample size is acceptable and independent selection is desirable.
- Use `sps` or `pareto` as fast high-entropy order-sampling alternatives when approximate first- and second-order quantities are acceptable.
- Use `cube` when balancing on auxiliary variables is more important than exact second-order inclusion probabilities.
- Use `lpm2` or `scps` when the study variable is spatially structured and a well-spread sample matters more than joint inclusion probabilities. LPM2 makes local pairwise competitions and SCPS distributes each decision over the nearest feasible neighbours. Variance is then usually estimated with local-neighbourhood estimators.

## Custom methods

`register_method()` lets you plug any unequal-probability, balanced, or
spatially balanced sampling algorithm into sondage's dispatchers and
generics. Methods registered with `type = "wor"` or `type = "wr"`
dispatch through `unequal_prob_wor()` / `unequal_prob_wr()`, and
balanced methods (`type = "balanced"`) through `balanced_wor()`, where
they declare which design inputs they use (`supports_aux`,
`supports_strata`, `supports_spread`):

Here `type = "balanced"` names the dispatcher family, which includes
spatially balanced designs. Spread-only methods such as LPM2 and SCPS do not
exactly balance auxiliary totals; their `supports_aux = FALSE` metadata makes
that distinction enforceable rather than silently ignoring `aux`.

```{r}
# A simple randomized pivotal sampler written in R
random_pivotal_sample <- function(pik, n = NULL, prn = NULL, ...) {
  p <- pik
  tol <- 1e-06
  active <- which(p > tol & p < 1 - tol)
  while (length(active) >= 2L) {
    ij <- sample(active, 2L)
    i <- ij[1L]
    j <- ij[2L]
    total <- p[i] + p[j]
    if (total < 1) {
      if (runif(1) < p[i] / total) p[c(i, j)] <- c(total, 0)
      else p[c(i, j)] <- c(0, total)
    } else {
      if (runif(1) < (1 - p[j]) / (2 - total)) {
        p[c(i, j)] <- c(1, total - 1)
      } else {
        p[c(i, j)] <- c(total - 1, 1)
      }
    }
    active <- which(p > tol & p < 1 - tol)
  }
  sort(which(p > 0.5))
}

register_method("random_pivotal", type = "wor", sample_fn = random_pivotal_sample)

pik <- inclusion_prob(1:8, n = 3)
s <- unequal_prob_wor(pik, method = "random_pivotal")
s

unregister_method("random_pivotal")
```

See `vignette("custom-methods")` for more examples, including a custom
balanced method with stratification support, a spatially balanced
method using `spread`, and how to provide a `joint_fn` for variance
estimation.

## Why not sampling?

The `sampling` package (Tillé and Matei) is the reference toolkit for survey sampling in `R`, and it is more comprehensive than `sondage`. Many of the algorithms here follow the methods it established, and `sondage` would not exist without it.

What `sondage` adds is speed. The sampling algorithms are written in `C`, so they usually scale better to large populations. Every sampling function also returns a design object with `S3` generics for inclusion probabilities, joint inclusion probabilities, and variance quantities, so the results are easy to carry into downstream work.

The two packages are complementary rather than competing. With `register_method()` you can plug any unequal probability algorithm from `sampling` into the `sondage` dispatchers and generics, so they can be used together.

## References

Brewer, K.R.W. and Donadio, M.E. (2003). The High Entropy Variance of the Horvitz-Thompson Estimator. *Survey Methodology*, 29(2), 189-196.

Chauvet, G. (2009). Stratified balanced sampling. *Survey Methodology*, 35, 115-119.

Chromy, J.R. (1979). Sequential sample selection methods. *Proceedings of the Survey Research Methods Section, American Statistical Association*, 401-406.

Chromy, J.R. (2009). Some generalizations of the Horvitz-Thompson estimator. *Proceedings of the Survey Research Methods Section, American Statistical Association*.

Deville, J.C. and Tillé, Y. (2004). Efficient balanced sampling: the cube method. *Biometrika*, 91(4), 893-912.

Grafström, A. (2012). Spatially correlated Poisson sampling. *Journal of Statistical Planning and Inference*, 142(1), 139-147.

Grafström, A., Lundström, N.L.P. and Schelin, L. (2012). Spatially balanced sampling through the pivotal method. *Biometrics*, 68(2), 514-520.

Tripet, A. and Tillé, Y. (2026). Balanced sampling with inequalities: application to category bounding, matrix rounding, and spread sampling. *Journal of the American Statistical Association*, 121(553), 796-806.

Tillé, Y. (2006). *Sampling Algorithms*. Springer.
