---
title: "Extending sondage with Custom Methods"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Extending sondage with Custom Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Overview

`sondage` ships with 12 built-in sampling methods, but researchers and agencies often need algorithms that are not included. The registration API lets you plug a custom unequal probability sampling method into the existing dispatchers and generics without modifying the package source.

After registering a method, it works everywhere a built-in method works: `unequal_prob_wor()`, `joint_inclusion_prob()`, `sampling_cov()`, batch mode (`nrep`), etc.

## The registration API

A single call to `register_method()` adds a new method:

```r
library(sondage)
register_method(
  name,                   # unique method name (character)
  type = "wor",           # "wor" or "wr"
  sample_fn,              # function(pik, n, prn, ...) -> integer indices
  joint_fn = NULL,        # function(pik, sample_idx, ...) -> matrix (optional)
  fixed_size = TRUE,      # does the method always return exactly n units?
  supports_prn = FALSE    # does the method support permanent random numbers?
)
```

Two contracts must be respected:

- **`sample_fn(pik, n = NULL, prn = NULL, ...)`** receives the inclusion probability vector and returns an integer vector of selected unit indices (1-based).
- **`joint_fn(pik, sample_idx = NULL, ...)`** (optional) returns an $N \times N$ joint inclusion probability matrix when `sample_idx` is `NULL`, or a submatrix for the given indices when it is not.

Helper functions `registered_methods()`, `is_registered_method()`, and `unregister_method()` manage the registry.

## Example 1: Sampford's method with high-entropy joint probabilities

Sampford's [-@sampford1967] rejection procedure draws exact $\pi$ps samples. The first unit is drawn with probability proportional to $\pi_k$; the remaining $n - 1$ units are drawn independently with replacement, each with probability proportional to $\pi_k / (1 -\pi_k)$. The draw is accepted only when all $n$ units are distinct.

```{r sampford-fn}
sampford_sample <- function(pik, n = NULL, prn = NULL, ...) {
  N <- length(pik)
  if (n == 1L) {
    return(sample.int(N, 1L, prob = pik))
  }
  q <- pik / (1 - pik)
  repeat {
    first <- sample.int(N, 1L, prob = pik)
    rest  <- sample.int(N, n - 1L, replace = TRUE, prob = q)
    s <- c(first, rest)
    if (anyDuplicated(s) == 0L)
      return(sort(s))
  }
}
```

The acceptance rate decreases as $n / N$ grows, so this implementation is best suited for moderate sampling fractions.

### Joint probabilities via `he_jip()`

Sampford's design is a high-entropy design, so the high-entropy (HE) approximation of @brewer2003 applies. Rather than reimplementing the formula, we use `he_jip()`, which `sondage` exports for exactly this purpose. It calls the same optimised C code used internally for Brewer, SPS, Pareto, and cube methods, and it already matches the `joint_fn` signature expected by `register_method()`.

For designs closer to conditional Poisson (rejective) sampling, `hajek_jip()` provides a lighter alternative based on @hajek1964.

### Registering and using Sampford's method

```{r sampford-register}
library(sondage)

register_method(
  "sampford",
  type       = "wor",
  sample_fn  = sampford_sample,
  joint_fn   = he_jip,
  fixed_size = TRUE
)
```

Once registered, the method is available through the standard
dispatcher:

```{r sampford-use}
pik <- inclusion_prob(c(2, 3, 4, 5, 6, 7, 8, 9), n = 4)
pik

s <- unequal_prob_wor(pik, method = "sampford")
s
```

All generics work transparently:

```{r sampford-generics}
# Joint inclusion probabilities
pikl <- joint_inclusion_prob(s)
round(pikl, 4)

# Sampling covariance (Sen-Yates-Grundy check quantities)
round(sampling_cov(s, weighted = TRUE), 4)

# sampled_only for large populations
joint_inclusion_prob(s, sampled_only = TRUE)
```

### Verifying with simulation

We can check that the empirical inclusion frequencies match the
target $\pi_k$ and that the empirical pairwise frequencies match
the HE approximation.

```{r sampford-verify}
sim <- unequal_prob_wor(pik, method = "sampford", nrep = 5000)
dim(sim$sample) # n x nrep

# First-order: empirical vs target
freq <- tabulate(sim$sample, nbins = length(pik)) / 5000
cbind(target = pik, empirical = freq)
```

```{r sampford-verify-joint}
# Second-order: empirical vs HE approximation
N <- length(pik)
co_occur <- matrix(0, N, N)
for (j in seq_len(5000)) {
  idx <- sim$sample[, j]
  co_occur[idx, idx] <- co_occur[idx, idx] + 1
}
diag(co_occur) <- 0
empirical_jip <- co_occur / 5000
diag(empirical_jip) <- freq

he_pikl <- he_jip(pik)

# Compare a few off-diagonal pairs
pairs <- data.frame(
  i      = c(1, 2, 3, 5),
  j      = c(8, 7, 6, 8),
  HE     = round(sapply(1:4, \(k) he_pikl[c(1,2,3,5)[k], c(8,7,6,8)[k]]), 4),
  empirical = round(sapply(1:4, \(k) empirical_jip[c(1,2,3,5)[k], c(8,7,6,8)[k]]), 4)
)
pairs
```

The HE approximation is close because Sampford's design is itself a high-entropy design.

```{r sampford-cleanup, include = FALSE}
unregister_method("sampford")
```


## Example 2: Wrapping an external package (sampling::UPtille)

When an algorithm is already implemented in another package, the
wrapper is minimal. Here we wrap `UPtille` and `UPtillepi2` from the
**sampling** package [@tille2006].

```{r tille, eval = requireNamespace("sampling", quietly = TRUE)}
tille_sample <- function(pik, n = NULL, prn = NULL, ...) {
  which(as.logical(sampling::UPtille(pik)))
}

tille_joint <- function(pik, sample_idx = NULL, ...) {
  pikl <- sampling::UPtillepi2(pik)
  if (!is.null(sample_idx)) {
    pikl <- pikl[sample_idx, sample_idx, drop = FALSE]
  }
  pikl
}

register_method(
  "tille",
  type       = "wor",
  sample_fn  = tille_sample,
  joint_fn   = tille_joint,
  fixed_size = TRUE
)

pik <- inclusion_prob(c(2, 3, 4, 5, 6, 7, 8, 9), n = 4)
s <- unequal_prob_wor(pik, method = "tille")
s

# Exact joint inclusion probabilities from UPtillepi2
round(joint_inclusion_prob(s), 4)

# Full variance estimation chain
round(sampling_cov(s), 4)
```

```{r tille-cleanup, include = FALSE, eval = requireNamespace("sampling", quietly = TRUE)}
unregister_method("tille")
```


## Writing a `joint_fn`

A `joint_fn` is optional but enables `joint_inclusion_prob()`,
`joint_expected_hits()`, and `sampling_cov()`. Four common
strategies:

1. **Exact formula.** When the design has a known closed form
   (e.g., `UPtillepi2` above).

2. **`he_jip()`.** The high-entropy approximation
   [@brewer2003]. Best for maximum-entropy and high-entropy
   designs (Sampford, Tille, and most $\pi$ps procedures). Uses
   optimised C code internally. Pass it directly:
   `joint_fn = he_jip`.

3. **`hajek_jip()`.** The Hajek [-@hajek1964] approximation
   based on conditional Poisson (rejective) sampling theory.
   Simpler formula and slightly cheaper, but generally a bit
   less accurate than `he_jip()`. Best when the design is
   obtained by conditioning independent Poisson trials on the
   sample size. Pass it directly: `joint_fn = hajek_jip`.

4. **Other approximations**. Other methods exist in the literature and can be
   implemented as custom `joint_fn` functions. @tille1996 reviews
   several alternatives, including approximations based on
   Poisson, rejective, and successive sampling theory, each with
   different accuracy--computation trade-offs. Any function that
   accepts `(pik, sample_idx = NULL, ...)` and returns a symmetric
   matrix is a valid `joint_fn`.

5. **Monte Carlo estimation.** Resample $B$ times and estimate
   $\hat{\pi}_{ij} = B^{-1} \sum_{b=1}^B I(i \in S_b)\, I(j \in S_b)$.
   Slow but universal:

```{r mc-joint, eval = FALSE}
mc_joint <- function(pik, sample_idx = NULL, ..., B = 5000) {
  N <- length(pik)
  n <- as.integer(round(sum(pik)))
  co <- matrix(0, N, N)
  for (b in seq_len(B)) {
    s <- my_sampler(pik, n = n)
    co[s, s] <- co[s, s] + 1
  }
  pikl <- co / B
  diag(pikl) <- tabulate(unlist(
    replicate(B, my_sampler(pik, n = n), simplify = FALSE)
  ), nbins = N) / B
  if (!is.null(sample_idx)) {
    pikl <- pikl[sample_idx, sample_idx, drop = FALSE]
  }
  pikl
}
```

The `sample_idx` argument enables the `sampled_only = TRUE` path
in `joint_inclusion_prob()`. When `sample_idx` is non-NULL, you
may either (a) compute the full matrix and subset as above, or
(b) skip rows/columns not in `sample_idx` for efficiency.

## Session persistence

The registry lives in the package namespace and resets when
`sondage` is reloaded. To make a registration persistent across
sessions, place the `register_method()` call in your `.Rprofile` or
in a project-level setup script.

## References
