---
title: "Extending sondage with Custom Methods"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Extending sondage with Custom Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Overview

`sondage` ships with 13 built-in sampling methods, but researchers and agencies often need algorithms that are not included. The registration API lets you plug a custom unequal probability or balanced sampling method into the existing dispatchers and generics without modifying the package source.

After registering a method, it works everywhere a built-in method works: `unequal_prob_wor()`, `unequal_prob_wr()`, `balanced_wor()`, `joint_inclusion_prob()`, `sampling_cov()`, batch mode (`nrep`), etc.

## The registration API

A single call to `register_method()` adds a new method:

```r
library(sondage)
register_method(
  name,                     # unique method name (character)
  type = "wor",             # "wor", "wr", or "balanced"
  sample_fn,                # function(pik, n, prn, ...) -> integer indices
  joint_fn = NULL,          # function(pik, sample_idx, ...) -> matrix (optional)
  fixed_size = TRUE,        # does the method always return exactly n units?
  variance_family = NULL,   # declared variance-estimation family (optional)
  supports_prn = FALSE,     # does the method support permanent random numbers?
  supports_aux = TRUE,      # balanced only: does it use balancing variables?
  supports_strata = FALSE,  # balanced only: does it handle stratification?
  supports_spread = FALSE   # balanced only: does it handle spatial spreading?
)
```

Two contracts must be respected:

- **`sample_fn(pik, n = NULL, prn = NULL, ...)`** receives the inclusion probability vector and returns an integer vector of selected unit indices (1-based). Balanced methods use the signature `sample_fn(pik, n = NULL, aux = NULL, ...)` instead, where `aux` is the auxiliary balancing matrix passed to `balanced_wor()` (or `NULL`).
- **`joint_fn(pik, sample_idx = NULL, ...)`** (optional) returns an $N \times N$ joint inclusion probability matrix when `sample_idx` is `NULL`, or a submatrix for the given indices when it is not.

Capability flags describe what the method can do, and the dispatchers enforce them. A `"wor"` method that sets `supports_prn = TRUE` receives the `prn` vector for sample coordination, while others get a warning that `prn` is ignored. A `"balanced"` method only receives a `strata` argument when registered with `supports_strata = TRUE`, and a `spread` argument (spatial coordinates) when registered with `supports_spread = TRUE`. Spread-only methods that never look at balancing variables should declare `supports_aux = FALSE`. In every case, passing a design input to a method without the matching capability is an error rather than a warning, because silently dropping `aux`, `strata`, or `spread` would change the design. The simplest balanced registration therefore needs nothing beyond `aux`.

`variance_family` declares how design-based variance should be estimated for samples drawn with the method — `"srs"`, `"pps_brewer"`, `"poisson"`, `"wr"`, or `"unsupported"`. `sondage` does not use it itself, but packages that export samples for variance estimation do: without a declaration they must infer a treatment from `type` and `fixed_size`, and for a random-size WOR method no safe inference exists — a Poisson-type method (independent selections) and a correlated random-size scheme need different estimators. The declaration is an assertion the author is responsible for; `?register_method` lists the constraints between `variance_family`, `type`, and `fixed_size`, and Example 1 shows how to check a declaration by simulation. When in doubt, declare `"unsupported"` rather than guess.

Helper functions `registered_methods()`, `is_registered_method()`, and `unregister_method()` manage the registry.

## Example 1: Sampford's method with high-entropy joint probabilities

Sampford's [-@sampford1967] rejection procedure draws exact $\pi$ps samples. The first unit is drawn with probability proportional to $\pi_k$; the remaining $n - 1$ units are drawn independently with replacement, each with probability proportional to $\pi_k / (1 -\pi_k)$. The draw is accepted only when all $n$ units are distinct.

```{r sampford-fn}
sampford_sample <- function(pik, n = NULL, prn = NULL, ...) {
  N <- length(pik)
  if (n == 1L) {
    return(sample.int(N, 1L, prob = pik))
  }
  q <- pik / (1 - pik)
  repeat {
    first <- sample.int(N, 1L, prob = pik)
    rest  <- sample.int(N, n - 1L, replace = TRUE, prob = q)
    s <- c(first, rest)
    if (anyDuplicated(s) == 0L)
      return(sort(s))
  }
}
```

The acceptance rate decreases as $n / N$ grows, so this implementation is best suited for moderate sampling fractions.

### Joint probabilities via `he_jip()`

Sampford's design is a high-entropy design, so the high-entropy (HE) approximation of @brewer2003 applies. Rather than reimplementing the formula, we use `he_jip()`, which `sondage` exports for exactly this purpose. It calls the same optimised C code used internally for Brewer, SPS, Pareto, and cube methods, and it already matches the `joint_fn` signature expected by `register_method()`.

For designs closer to conditional Poisson (rejective) sampling, `hajek_jip()` provides a lighter alternative based on @hajek1964.

### Registering and using Sampford's method

```{r sampford-register}
library(sondage)

register_method(
  "sampford",
  type            = "wor",
  sample_fn       = sampford_sample,
  joint_fn        = he_jip,
  fixed_size      = TRUE,
  variance_family = "pps_brewer"
)
```

Sampford's design is fixed-size unequal-probability WOR, so
`"pps_brewer"` is the natural `variance_family` declaration:
downstream variance estimation treats it like the built-in fixed-size
$\pi$ps methods (Brewer, SPS, Pareto). The declaration is verified by
simulation below.

Once registered, the method is available through the standard
dispatcher:

```{r sampford-use}
pik <- inclusion_prob(c(2, 3, 4, 5, 6, 7, 8, 9), n = 4)
pik

s <- unequal_prob_wor(pik, method = "sampford")
s
```

All generics work transparently:

```{r sampford-generics}
# Joint inclusion probabilities
pikl <- joint_inclusion_prob(s)
round(pikl, 4)

# Sampling covariance (Sen-Yates-Grundy check quantities)
round(sampling_cov(s, weighted = TRUE), 4)

# sampled_only for large populations
joint_inclusion_prob(s, sampled_only = TRUE)
```

### Verifying with simulation

We can check that the empirical inclusion frequencies match the
target $\pi_k$ and that the empirical pairwise frequencies match
the HE approximation.

```{r sampford-verify}
sim <- unequal_prob_wor(pik, method = "sampford", nrep = 5000)
dim(sim$sample) # n x nrep

# First-order: empirical vs target
freq <- tabulate(sim$sample, nbins = length(pik)) / 5000
cbind(target = pik, empirical = freq)
```

```{r sampford-verify-joint}
# Second-order: empirical vs HE approximation
N <- length(pik)
co_occur <- matrix(0, N, N)
for (j in seq_len(5000)) {
  idx <- sim$sample[, j]
  co_occur[idx, idx] <- co_occur[idx, idx] + 1
}
diag(co_occur) <- 0
empirical_jip <- co_occur / 5000
diag(empirical_jip) <- freq

he_pikl <- he_jip(pik)

# Compare a few off-diagonal pairs
pairs <- data.frame(
  i      = c(1, 2, 3, 5),
  j      = c(8, 7, 6, 8),
  HE     = round(sapply(1:4, \(k) he_pikl[c(1,2,3,5)[k], c(8,7,6,8)[k]]), 4),
  empirical = round(sapply(1:4, \(k) empirical_jip[c(1,2,3,5)[k], c(8,7,6,8)[k]]), 4)
)
pairs
```

The HE approximation is close because Sampford's design is itself a high-entropy design.

The `variance_family` declaration can be checked the same way.
`"pps_brewer"` asserts that the variance of a Horvitz-Thompson
estimator under this design is described by the joint probabilities
implied by the marginal $\pi_k$. Comparing the Monte Carlo variance of
the HT total for a study variable against the variance implied by the
declared family,
$V(\hat{t}_y) = \sum_k \sum_l (\pi_{kl} - \pi_k \pi_l)
\frac{y_k}{\pi_k} \frac{y_l}{\pi_l}$:

```{r sampford-verify-variance}
y <- c(5, 13, 8, 22, 16, 34, 27, 41) # a study variable
ht <- apply(sim$sample, 2, function(idx) sum(y[idx] / pik[idx]))

Delta <- he_pikl - outer(pik, pik) # diagonal is pik * (1 - pik)
yp <- y / pik
c(monte_carlo = var(ht), declared_family = sum(Delta * outer(yp, yp)))
```

The Monte Carlo value is the truth, up to simulation error; the
declared-family value inherits the HE approximation. The gap is
noticeably larger on the variance scale than the joint-probability
comparison above suggests, because $\Delta_{kl} = \pi_{kl} - \pi_k
\pi_l$ is a small difference of close quantities: percent-level
errors in $\pi_{kl}$ are amplified, and $n/N = 1/2$ is a demanding
sampling fraction for any high-entropy approximation. The exact
Sampford joint probabilities confirm that the discrepancy is
approximation error, not a wrong declaration:

```{r sampford-verify-exact, eval = requireNamespace("sampling", quietly = TRUE)}
pikl_exact <- sampling::UPsampfordpi2(pik)
Delta_exact <- pikl_exact - outer(pik, pik)
diag(Delta_exact) <- pik * (1 - pik)
sum(Delta_exact * outer(yp, yp)) # next to the Monte Carlo value
```

This is the resolution of the check: it validates the order of
magnitude, which is what a family declaration asserts. A misdeclared
family is wrong by orders of magnitude, not by tens of percent — a
random-size method with correlated selections declared `"poisson"`,
or a Poisson-type method treated as fixed-size $\pi$ps, can report
standard errors near zero. When no family fits, declare
`"unsupported"` instead: an error at export time is recoverable, a
silently wrong standard error is not.

```{r sampford-cleanup, include = FALSE}
unregister_method("sampford")
```


## Example 2: Wrapping an external package (sampling::UPtille)

When an algorithm is already implemented in another package, the
wrapper is minimal. Here we wrap `UPtille` and `UPtillepi2` from the
**sampling** package [@tille2006].

```{r tille, eval = requireNamespace("sampling", quietly = TRUE)}
tille_sample <- function(pik, n = NULL, prn = NULL, ...) {
  which(as.logical(sampling::UPtille(pik)))
}

tille_joint <- function(pik, sample_idx = NULL, ...) {
  pikl <- sampling::UPtillepi2(pik)
  if (!is.null(sample_idx)) {
    pikl <- pikl[sample_idx, sample_idx, drop = FALSE]
  }
  pikl
}

register_method(
  "tille",
  type       = "wor",
  sample_fn  = tille_sample,
  joint_fn   = tille_joint,
  fixed_size = TRUE
)

pik <- inclusion_prob(c(2, 3, 4, 5, 6, 7, 8, 9), n = 4)
s <- unequal_prob_wor(pik, method = "tille")
s

# Exact joint inclusion probabilities from UPtillepi2
round(joint_inclusion_prob(s), 4)

# Full variance estimation chain
round(sampling_cov(s), 4)
```

```{r tille-cleanup, include = FALSE, eval = requireNamespace("sampling", quietly = TRUE)}
unregister_method("tille")
```

## Example 3: A custom balanced method

Balanced methods register with `type = "balanced"` and dispatch
through `balanced_wor()`. The minimal contract only involves `aux`,
so wrapping an aux-only algorithm is a one-liner. Here we wrap the
cube implementation from the **sampling** package, whose landing
phase uses linear programming and can therefore give different
samples than the built-in `"cube"` method.

Note two conventions. The wrapper receives `aux` exactly as the
caller supplied it (validated, but without the sample-size
constraint prepended), so we `cbind(pik, aux)` ourselves because
`samplecube()` expects the size constraint as a balancing column.
And since the cube method produces a high-entropy design, we can
pass `he_jip` as the `joint_fn`, which is exactly what the built-in
method uses.

```{r cube-lp, eval = requireNamespace("sampling", quietly = TRUE)}
cube_lp_sample <- function(pik, n = NULL, aux = NULL, ...) {
  X <- cbind(pik, aux)
  which(sampling::samplecube(X, pik, comment = FALSE) == 1)
}

register_method(
  "cube_lp",
  type      = "balanced",
  sample_fn = cube_lp_sample,
  joint_fn  = he_jip
)

pik <- inclusion_prob(c(2, 3, 4, 5, 6, 7, 8, 9), n = 4)
x <- matrix(c(10, 20, 15, 25, 30, 35, 40, 45))
s <- balanced_wor(pik, aux = x, method = "cube_lp")
s

# Balancing check: HT estimate of the aux total vs the true total
colSums(x[s$sample, , drop = FALSE] / pik[s$sample]) - colSums(x)
```

To support stratification as well, register with
`supports_strata = TRUE` and add a `strata` argument to the
sampler. The dispatcher passes `strata` as dense integer labels
`1:H` (only when the caller supplies them), and demotes
`fixed_size` with a warning when per-stratum `sum(pik)` is not
close to an integer, mirroring the built-in method.

```{r cube-lp-strata, eval = requireNamespace("sampling", quietly = TRUE)}
cube_lp_stratified <- function(pik, n = NULL, aux = NULL, strata = NULL, ...) {
  if (is.null(strata)) {
    return(cube_lp_sample(pik, n = n, aux = aux))
  }
  X <- if (is.null(aux)) matrix(pik, ncol = 1) else cbind(pik, aux)
  which(sampling::balancedstratification(X, strata, pik, comment = FALSE) == 1)
}

register_method(
  "cube_lp_str",
  type            = "balanced",
  sample_fn       = cube_lp_stratified,
  joint_fn        = he_jip,
  supports_strata = TRUE
)

pik <- rep(0.5, 8)
strata <- rep(1:2, each = 4)
s <- balanced_wor(pik, aux = matrix(as.double(1:8)), strata = strata,
                  method = "cube_lp_str")

# Within-stratum sample sizes are preserved
tabulate(strata[s$sample], nbins = 2)
```

```{r cube-lp-cleanup, include = FALSE, eval = requireNamespace("sampling", quietly = TRUE)}
unregister_method("cube_lp")
unregister_method("cube_lp_str")
```

## Example 4: A spatial method with `spread`

Spatially balanced (well-spread) designs select units that are far
apart in space, which improves precision whenever the study variable
is spatially structured. Registered balanced methods opt into
spatial spreading with `supports_spread = TRUE`, and then receive
the coordinate matrix that the caller passes to
`balanced_wor(spread = )`.

Here is a compact pure-R implementation of the local pivotal method
[@grafstrom2012], which repeatedly lets a random undecided unit and
its nearest undecided neighbour compete in a pivotal step. Packages
such as **BalancedSampling** or **rsamplr** provide fast compiled
implementations of this and related algorithms (SCPS, local cube),
and wrapping one of those instead only changes the function body.

```{r lpm2-fn}
lpm2_sample <- function(pik, n = NULL, aux = NULL, spread = NULL, ...) {
  d <- as.matrix(dist(spread))
  diag(d) <- Inf
  p <- pik
  eps <- 1e-9
  repeat {
    u <- which(p > eps & p < 1 - eps)
    if (length(u) == 0L) {
      break
    }
    if (length(u) == 1L) {
      p[u] <- as.numeric(runif(1) < p[u])
      break
    }
    i <- u[sample.int(length(u), 1L)]
    v <- u[u != i]
    j <- v[which.min(d[i, v])]
    s <- p[i] + p[j]
    if (s > 1) {
      if (runif(1) < (1 - p[j]) / (2 - s)) {
        p[i] <- 1
        p[j] <- s - 1
      } else {
        p[j] <- 1
        p[i] <- s - 1
      }
    } else {
      if (runif(1) < p[j] / s) {
        p[j] <- s
        p[i] <- 0
      } else {
        p[i] <- s
        p[j] <- 0
      }
    }
  }
  which(p > 1 - eps)
}

register_method(
  "lpm2",
  type            = "balanced",
  sample_fn       = lpm2_sample,
  supports_aux    = FALSE,
  supports_spread = TRUE
)
```

Since the local pivotal method spreads but never balances on
auxiliary variables, it is registered with `supports_aux = FALSE`.
A caller who passes `aux` then gets an immediate error instead of a
sample that silently ignored the requested balancing constraints.

```{r lpm2-run}
set.seed(42)
N <- 200
coords <- cbind(runif(N), runif(N))
pik <- rep(0.15, N)

s <- balanced_wor(pik, spread = coords, method = "lpm2")
s

# Spread diagnostic: mean distance to the nearest sampled neighbour
# (larger is better spread)
nn_dist <- function(idx) {
  d <- as.matrix(dist(coords[idx, ]))
  diag(d) <- Inf
  mean(apply(d, 1, min))
}
c(lpm2 = nn_dist(s$sample), srs = nn_dist(sample.int(N, s$n)))
```

A note on variance estimation. Well-spread designs have no
tractable joint inclusion probabilities (the high-entropy
approximation does not apply, since spreading deliberately drives
nearby joint probabilities toward zero), so `joint_fn` is usually
left as `NULL` and `joint_inclusion_prob()` / `sampling_cov()`
will error for these methods. In practice, variance for spatially
balanced samples is estimated with local-neighbourhood estimators
such as the local mean estimator of @grafstrom2014, which sondage
does not implement yet.

```{r lpm2-cleanup, include = FALSE}
unregister_method("lpm2")
```

## Writing a `joint_fn`

A `joint_fn` is optional but enables `joint_inclusion_prob()`,
`joint_expected_hits()`, and `sampling_cov()`. Four common
strategies:

1. **Exact formula.** When the design has a known closed form
   (e.g., `UPtillepi2` above).

2. **`he_jip()`.** The high-entropy approximation
   [@brewer2003]. Best for maximum-entropy and high-entropy
   designs (Sampford, Tillé, and most $\pi$ps procedures). Uses
   optimised C code internally. Pass it directly:
   `joint_fn = he_jip`.

3. **`hajek_jip()`.** The Hajek [-@hajek1964] approximation
   based on conditional Poisson (rejective) sampling theory.
   Simpler formula and slightly cheaper, but generally a bit
   less accurate than `he_jip()`. Best when the design is
   obtained by conditioning independent Poisson trials on the
   sample size. Pass it directly: `joint_fn = hajek_jip`.

4. **Other approximations**. Other methods exist in the literature and can be
   implemented as custom `joint_fn` functions. @tille1996 reviews
   several alternatives, including approximations based on
   Poisson, rejective, and successive sampling theory, each with
   different accuracy--computation trade-offs. Any function that
   accepts `(pik, sample_idx = NULL, ...)` and returns a symmetric
   matrix is a valid `joint_fn`.

5. **Monte Carlo estimation.** Resample $B$ times and estimate
   $\hat{\pi}_{ij} = B^{-1} \sum_{b=1}^B I(i \in S_b)\, I(j \in S_b)$.
   Slow but universal:

```{r mc-joint, eval = FALSE}
mc_joint <- function(pik, sample_idx = NULL, ..., B = 5000) {
  N <- length(pik)
  n <- as.integer(round(sum(pik)))
  co <- matrix(0, N, N)
  for (b in seq_len(B)) {
    s <- my_sampler(pik, n = n)
    co[s, s] <- co[s, s] + 1
  }
  pikl <- co / B
  diag(pikl) <- tabulate(unlist(
    replicate(B, my_sampler(pik, n = n), simplify = FALSE)
  ), nbins = N) / B
  if (!is.null(sample_idx)) {
    pikl <- pikl[sample_idx, sample_idx, drop = FALSE]
  }
  pikl
}
```

The `sample_idx` argument enables the `sampled_only = TRUE` path
in `joint_inclusion_prob()`. When `sample_idx` is non-NULL, you
may either (a) compute the full matrix and subset as above, or
(b) skip rows/columns not in `sample_idx` for efficiency.

## Session persistence

The registry lives in the package namespace and resets when
`sondage` is reloaded. To make a registration persistent across
sessions, place the `register_method()` call in your `.Rprofile` or
in a project-level setup script.

## References
